---
title: "Data Visualization in R"
author: "Edgar Orong"
date: "2023-07-22"
output:
  html_document: default
  pdf_document: default
---

## Setting up my environment
Notes: Setting up my R environment by loading the 'ggplot2' and 'palmerpenguins' packages

```{r loading packages}
library(ggplot2)
library("palmerpenguins")
```

## Core concepts of ggplot2
1. Aesthetic. A visual property of an object in your plot
2. Geoms. The geometric object used to represent your data
3. Facets. Let's you display smaller groups or subsets of your data
4. Labels and annotations. Let you customize your plot


## Three steps of creating a plot
1. Start with the ggplot function and choose a dataset to work with
2. Add a geom_function to display your data
3. Map the variables you want to plot in the arguments of the aes() function


**Example 1:**
```{r Steps of creating a plot, warning=FALSE}
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g))
```

**Example 2:**
```{r another example, warning=FALSE}
ggplot(data = penguins) +
  geom_point(mapping = aes(x = bill_length_mm, y = bill_depth_mm))
```
## Changing Aesthetic

### 1. Changing color for every variable
This will map colors into specific variable (species)

```{r changing color for every species, warning=FALSE}
ggplot(data = penguins) +
    geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species))
```

### 2. Changing shape for every variable
This will map different shape for every variable (species)

```{r changing shape for every variable, warning=FALSE}
 ggplot(data = penguins) +
    geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, shape = species, color = species))
```

### 3. Adding size for every variable

```{r Adding size, warning=FALSE}
ggplot(data = penguins) +
    geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species, size = species))
  
```

### 4. Alpha aesthetics
This controls the transparency of the points. Alpha is a good option when you got a dense plot with lots of data points

```{r alpha aesthetics, warning=FALSE}
  ggplot(data = penguins) +
    geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, alpha = species))
```

### 5. To change color but not tie or map a color to a specific variable
**Note:** if we want to change the overall appearance of our plot without regard specific variable, we write code outside of the aes() function.

```{r changing color but not tie or map a color to a specific variable, warning=FALSE}
 ggplot(data = penguins) +
    geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, alpha = species), color = "blue")
```

## Different GEOM functions to create different types of plots

### 1. Geom_point

```{r geom_point, warning=FALSE}
  ggplot(data = penguins) + 
    geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g))
```

### 2. Geom_bar
**Note:** R automatically counts how many times x value appears in the data and then shows the count on the y axis.

```{r geom_bar}
 ggplot(data = diamonds) +
    geom_bar(mapping = aes(x = cut))
```

#### To outline geom_bar with color
      
```{r geom_bar with outline color}
  ggplot(data = diamonds) +
            geom_bar(mapping = aes(x = cut, color = cut))
```

#### To fill bar with color

```{r geom_bar with color fill}
 ggplot(data = diamonds) +
            geom_bar(mapping = aes(x = cut, fill = cut))
```

#### To make stacked bar, map fill with other variable other than cut

```{r stacked_bar}
      ggplot(data = diamonds) +
            geom_bar(mapping = aes(x = cut, fill = clarity))
```
      
### 3. Geom_line

```{r geom_line, warning=FALSE}
  ggplot(data = penguins) +
    geom_line(mapping = aes(x = flipper_length_mm, y = body_mass_g))
```

#### geom_smooth
**Note:** geom_smooth is useful for showing general trends in our data.

```{r geom_smooth, message=FALSE, warning=FALSE}
ggplot(data = penguins) +
  geom_smooth(mapping = aes(x = flipper_length_mm, y = body_mass_g))
```

#### Linetype
**Note:** To plot separate line for each species of penguins, add linetype aesthetic to the code

```{r plot separate line, message=FALSE, warning=FALSE}
  ggplot(data = penguins) +
          geom_smooth(mapping = aes(x = flipper_length_mm, y = body_mass_g, linetype = species))
```

#### Combining two geom functions

```{r combining two geom functions, message=FALSE, warning=FALSE}
 ggplot(data = penguins) +
          geom_smooth(mapping = aes(x = flipper_length_mm, y = body_mass_g)) +
          geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g), color = "black")
```

### 4. Geom_jitter
Creates a scatter plot and then adds a small amount of random noise to each point in the plot.

**Note**: Jittering helps us deal with overplotting which happens when the datapoints in a plot overlap with each other. It makes the points easier to find

```{r geom_jitter, message=FALSE, warning=FALSE}
  ggplot(data = penguins) +
    geom_jitter(mapping = aes(x = flipper_length_mm, y = body_mass_g))
```


## Facets
Lets you display smaller groups, or subsets, of your data. Faceting can help you discover new patterns in your data and focus on relationships between different variables. 

### Two facet functions

## 1. Facet_wrap
Is used to facet your plot by a SINGLE VARIABLE

**Example 1:**

```{r facet_wrap 1, message=FALSE, warning=FALSE}
ggplot(data = penguins) +
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  facet_wrap(~species)
```

**Example 2:**

```{r facet_wrap 2, message=FALSE, warning=FALSE}
      ggplot(data = diamonds) +
        geom_bar(mapping = aes(x = color, fill = cut)) +
        facet_wrap(~cut)
```

## 2. Facet_grid
Is used to facet your plot with TWO VARIABLES
**Note:** R facets vertically by the values of the first variable and horizontally by the values of second variable

```{r facet_grid, message=FALSE, warning=FALSE}
 ggplot(data = penguins) +
        geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
        facet_grid(sex~species)
```


## Labels and Annotations

**Labels:** Are titles, subtitles and captions that we put OUTSIDE OF THE GRID of our plot to indicate important information.



#### To add title
```{r to add title, message=FALSE, warning=FALSE}
ggplot(data = penguins) +
        geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
        labs(title = "Palmer Penguins: Body Mass vs. Flipper Length")
```



#### To add subtitle
```{r to add subtitle, message=FALSE, warning=FALSE}
ggplot(data = penguins) +
        geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
        labs(title = "Palmer Penguins: Body Mass vs. Flipper Length", subtitle = "Sample of Three Penguin Species")
```



#### To add caption
```{r to add caption, message=FALSE, warning=FALSE}
 ggplot(data = penguins) +
        geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
        labs(title = "Palmer Penguins: Body Mass vs. Flipper Length", subtitle = "Sample of Three Penguin Species", caption = "Data collected by Dr. Kristen Gorman" )
      
```



**Annotations:** Is used to put texts INSIDE THE GRID to explain or comment upon specific data points. In ggplot2, it can help explain the plot's purpose or highlight important data.


#### To put annotations inside the grid
```{r to put annotations, message=FALSE, warning=FALSE}
ggplot(data = penguins) +
        geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species))+
        labs(title = "Palmer Penguins: Body Mass vs. FLipper Length", subtitle = "Sample of Three Penguin Species", caption = "Data collected by Dr. Kristen Gorman") +
        annotate("text", x = 220, y = 3500, label = "The Gentoos are the largest")
```

#### To customize annotations
```{r to customize annotations, message=FALSE, warning=FALSE}
 ggplot(data = penguins) +
        geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species))+
        labs(title = "Palmer Penguins: Body Mass vs. FLipper Length", subtitle = "Sample of Three Penguin Species", caption = "Data collected by Dr. Kristen Gorman") +
        annotate("text", x = 220, y = 3500, label = "The Gentoos are the largest", color = "blue", fontface = "bold", size = 3, angle = 15)
```


### **PRO-TIP:** if you want to use less code, you can store your plot code in a variable and just add an annotation to it

```{r pro tip, message=FALSE, warning=FALSE}
p <-  ggplot(data = penguins) +
        geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species))+
        labs(title = "Palmer Penguins: Body Mass vs. FLipper Length", subtitle = "Sample of Three Penguin Species", caption = "Data collected by Dr. Kristen Gorman")
```

#### Now we can just add the assigned variable + the annotation

```{r P + annotation, message=FALSE, warning=FALSE}
p+ annotate("text", x = 220, y = 3500, label = "The Gentoos are the largest", color = "blue", fontface = "bold", size = 3, angle = 15)
```



## Saving your plot

**Approach 1:** Export option in the plots tab
      
**Approach 2:** ggsave() function

```{r approach 2, message=FALSE, warning=FALSE}
ggsave("Three Penguin Species.png")
```