---
title: 'IBM: R for Data Analysis (1)'
author: "Edgar Orong"
date: "2023-05-13"
output: html_document
---


## VECTOR AND FACTORS

**Vector** - one-dimensional array of objects and is a simple tool to store your data.

#### To create a vector:
**Type 1:** Numeric vector (to hold a list of numbers)
```{r}
 c(1,2,3,4,5,6,7,8,9,10) # Vector that contains the numbers from 1 to 10
 c(1:10) # Also vector that contains the numbers from 1 to 10
 c(10:1) # If you want the numbers in decreasing order from 10 to 1 
 c(1985, 1999, 2010, 2002)
```

Vectors can also be assigned to a variable
```{r}
  ratings <- c(81, 125) # vectors can also be assigned in a variable
  ratings / 60
```

**Type 2:** Character vector (to hold a list of strings)
```{r}
  c("Toy Story", "Akira", "The Artist", "City of God", "Toy Story", "Akira", "The Artist", "City of God")
```

**Type 3:** Logical vector (True or False)
```{r}
  movie_ratings <- c(7.3, 8.5, 8.3, 7.5, 6.9, 5.2, 8.2, 8.0, 7.9, 9.3)
  movie_ratings > 7.5 # marks True or False the values that meet the condition
```


**Factors** - are variables that can take an unlimited number of values. These variables are often referred to as 'categorical variables'.

#### To create a factor: 
**Step 1:** Create a vector first
```{r}
genre_vector <- c("Comedy", "Comedy", "Animation", "Animation", "Crime")
```

**Step 2:** Create factor
```{r}
genre_factor <- factor(genre_vector)
genre_factor
```

Levels in output corresponds to the three unique categories of the vector: Animation, comedy, crime


#### Summarizing factors:
```{r}
summary(genre_factor) # Shows the count or number of occurences of each of the component categories
```

      
#### Types of Factors (categorical variables)

**1. Nominal categorical variables** - have no implied order. Example is the genre_vector.
**2. Ordinal categorical variables** - have ordering. 
      
**To create ordinal variable:**
      
**Step 1:** Create a vector first
```{r}
movielength_vector <- c("Very short", "Short", "Medium", "Short", "Long", "Very short", "Very long")
```

**Step 2:** Create factor
```{r}
movielength_factor <- factor(movielength_vector, 
                                ordered = TRUE,
                                levels = c("Very short", "Short", "Medium", "Long", "Very long"))
movielength_factor
```

          
#### VECTOR OPERATIONS
          
**To NAME the elements of the vector, use the names() function**

example: We want to name the year vector with the title of the movie
```{r}
year <- c(1985, 1999, 2010, 2002)
names(year) <- c("The Breakfast Club", "American Beauty", "Black Swan", "Chicago")
```

**To call in the year vector using the name assigned:**
```{r}
year["Black Swan"]
```

**To return the LENGTH of a vector, use length() function**
```{r}
year <- c(1985, 1999, 2010, 2002)
length(year)
```

**To SORT the vector, use the sort() function**
```{r}
year <- c(1985, 1999, 2010, 2002)
      names(year) <- c("The Breakfast Club", "American Beauty", "Black Swan", "Chicago")
year_sorted <- sort(year)
year_sorted
```

**To find the smallest/largest number, use the min() OR max() function**
```{r}
year <- c(1985, 1999, 2010, 2002)
min(year)
```

```{r}
year <- c(1985, 1999, 2010, 2002)
max(year)
```

**To find the AVERAGE of numbers, use sum() / no. of elements function OR mean() function**
```{r}
cost_2014 <- c(8.6, 8.5, 8.1) #sum() function
sum(cost_2014) / 3
```

```{r}
cost_2014 <- c(8.6, 8.5, 8.1) #mean() function
mean(cost_2014)
```

**To summarize vector, use summary() function**
```{r}
cost_2014 <- c(8.6, 8.5, 8.1)
summary(cost_2014)
      
```

      
      
#### VECTOR INDEX
      
**To return a single value**
```{r}
cost_2014 <- c(8.6, 8.5, 8.1)
cost_2014[2] # to return the 2nd value in the vector
```

**To return multiple values**
```{r}
cost_2014 <- c(8.6, 8.5, 8.1)
cost_2014[c(2,3)] # returns the 1st and 2nd value in vector
```

**To return a range of values**
```{r}
cost_2014 <- c(8.6, 8.5, 8.1)
cost_2014[1:3]
```

**To return negative index values (to remove particular element from the output)**
```{r}
titles <- c("Black Swan", "Casino", "City of God", "Jumanji", "Toy Story")
titles[-1] #removes first index aka "Black Swan"
      
```

**Out of range index**
```{r}
titles <- c("Black Swan", "Casino", "City of God", "Jumanji", "Toy Story")
titles[8] # This will return an NA since there are only 5 index in the vector
```

**Conditional index**
```{r}
cost_2014 <- c(8.6, 8.5, 8.1)
cost_2014 > 8.3
```

**To retrieve the elements that are true:**
```{r}
cost_2014[cost_2014 > 8.3]
```

**Missing values. Use NA for missing values**
```{r}
age_restriction <- c(14, 12, 10, NA, 18, NA)
```


#### VECTOR ARITHMETIC
        
example 1
```{r}
age_restriction <- c(14, 16, 12, 10, 18, 18)
sequences <- c(2, 3, 0, 2, 6, 3)
multiply <- age_restriction * sequences
multiply
```

#example 2
```{r}
cost_2014 <- c(8.6, 8.5, 8.1)
multiply2 <- cost_2014 * 10
multiply2
```

  
  
## LISTS 

**Lists** - collections of objects like a vector but unlike a vector, the elements inside can differ in data type.

**To create**
```{r}
movie <- list("Toy Story", 1995, c("Animation", "Adventure", "Comedy"))
```

**To access**
```{r}
movie[2] #approach 1
movie[2:3] #to access a range
```

      
#### Named List
    
**To create**
```{r}
movie <- list(name = "Toy Story",
                    year = 1995,
                    genre = c("Animation", "Adventure", "Comedy"))
```

**To access**
```{r}
movie$genre # Approach 1 (listname+$+name)
movie["genre"] # Approach 2 (enclosed in a bracket)
```

**Adding items to the list**
```{r}
movie["age"] <- 5 # Simply add the item in square bracket then assignment operator then value. 
```

**To modify the value of the list**
```{r}
movie["age"] <- 6
```

**To remove elements from the list**
```{r}
movie["age"] <- NULL # Just assign a null value to the element
```

  

## ARRAY AND MATRICES 

### Array

**Arrays** - A single or multidimensional structure containing data of the the same data type.
      
**To create:**

**Step 1:** Create a vector first
```{r}
movie_vector <- c("Akira", "Toy Story", "Room", "The Wave", "Whiplash", "Star Wars", "The Ring", "The Artist", "Jumanji")
```

**Step 2:** Use the array() function
```{r}
movie_array <- array(movie_vector, dim = c(4,3)) # add dim (dimension) to define the no. of rows and columns
```


**To access an element in array:**
```{r}
movie_array[1,2] # name_of_array[row, column]
```


**To get the dimension of the array, use the dim() function**
```{r}
dim(movie_array)
```


**We can also do math in arrays**
```{r}
length_vector <- c(125, 81, 118, 81, 106, 121, 95, 100, 104)
length_array <- array(length_vector, dim = c(3,3))
length_array
```

**Let's add 5 to the array, to account for a 5-min bathroom break:**
```{r}
length_array + 5
```

       
        
#### Using Logical Conditions to Subset Arrays
        
**Which movies can I finish watching in two hours?**

Using a logical condition, we can check which movies are less than 2 hours long.
```{r}
mask_array <- length_array < 120
mask_array
```

Using this array of TRUEs and FALSEs, we can subset the array of movie names:
```{r}
x_vector <- c("Akira", "Toy Story", "Room", "The Wave", "Whiplash",
                      "Star Wars", "The Ring", "The Artist", "Jumanji")
x_array <- array(x_vector, dim = c(3,3))
x_array[mask_array]
```

        
**Find all movie titles with length less than or equal to 90 minutes**
```{r}
mask_array <- length_array <= 90
x_array[mask_array]
```

       
### Matrices

**Matrices** - Same with arrays but must be two-dimensional
        
**To create:**

**Step 1:** Create a vector first
```{r}
movie_vector <- c("Akira", "Toy Story", "Room", "The Wave", "Whiplash", "Star Wars", "The Ring", "The Artist", "Jumanji")
```

**Step 2:** Use matrix() function with three arguments (vector, nrow, ncol)
```{r}
movie_matrix <- matrix(movie_vector, nrow = 3, ncol = 3)
```

**By default, matrix arranges by column, to arrange in rows use byrow = TRUE as the 4th argument.**
```{r}
movie_matrix <- matrix(movie_vector, nrow = 3, ncol = 3, byrow = TRUE)
```

**To Access:**
```{r}
movie_matrix[2:3, 1:2] # extract elements in rows 2 to 3, and columns 1 to 2
movie_matrix[1,1]  # To retrieve "Akira", you should use [1,1] as it lies in the first row and first column.
```



      
## Concatenation

**Concatenation Function** - is used to combine two vectors into one vector. It combines values of both vectors. 
          
Lets create a new vector for the upcoming movies as a upcoming_movie variable and add them to the movie_vector variable to create a new_vector of movies.
```{r}
upcoming_movie <- c("Fast and Furious 8", "xXx: Return of Xander Cage", "Suicide Squad")
new_vector <- c(movie_vector, upcoming_movie)
new_vector

```

        
## Data Frames

**Data Frames** - A type of structure that contains correlated information. (ex. movie titles along with their corresponding years.)
    
**To create:**
```{r}
movies <- data.frame(name = c("Toy Story", "Akira", "The Breakfast Club", "The Artist", "Modern Times", "Fight Club", "City of God", "The Untouchables"), 
                             year = c(1995, 1998, 1985, 2011, 1936, 1999, 2002, 1987))
movies
```

        
**To access:**
```{r}
movies$name #(name of data frame + $ + column name)
movies[1] # use column number to access the column
movies[1,2] 
```

**To get information about the data frame structure**
```{r}
str(movies)
```

**Head and tails**
```{r}
head(movies) #will display the first 6 elements of the data frame
tail(movies) #will display last 6 elements of the data frame
```

**To insert a new column**
```{r}
movies['length'] <- c(81, 125, 97, 100, 87, 139, 130, 119)
```

**To insert a new row**
```{r}
movies <- rbind(movies, c(name = "Dr. Strangelove", 
                                   year = 1964,
                                   length = 94))
```

**To delete rows, use negative [-]**
```{r}
movies <- movies[-9,]
```

**To delete column, just assign a NULL value**
```{r}
movies["length"] <- NULL
```

**Now, let's try to get the row with name "Toy Story"**
         
**Find the rows with name "Toy Story"**
```{r}
selected <- movies["name"] == "Toy Story"
```

**Get the selected row(s) with 'name' and 'year' columns**
```{r}
toy_story <- movies[selected, c("name", "year")]
toy_story
```

         
        
         
        
        
        
        
        
        
        
        
        
        
        
    
    


