---
title: "Importing Data in R"
author: "Edgar Orong"
date: "2023-05-09"
output: html_document
---


## Importing Data in R


####  Download datasets

**CSV file**
```{r}
download.file("https://cf-courses-data.s3.us.cloud-object-storage.appdomain.cloud/IBMDeveloperSkillsNetwork-RP0101EN-Coursera/v2/dataset/movies-db.csv", 
              destfile="movies-db.csv")

```


**XLS file**

```{r}
download.file("https://cf-courses-data.s3.us.cloud-object-storage.appdomain.cloud/IBMDeveloperSkillsNetwork-RP0101EN-Coursera/v2/dataset/movies-db.xls", 
              destfile="movies-db.xls")

```


**To read CSV files into R, we use the core function read.csv.**

```{r}
my_data <- read.csv("movies-db.csv")
my_data
```


The data was loaded into the my_data variable. But instead of viewing all the data at once, we can use the head function to take a look at only the top six rows of our table, like so:

**Print out the first six rows of my_data**

```{r}
head(my_data)

```


Additionally, you may want to take a look at the structure of your newly created table. R provides us with a function that summarizes an entire table's properties, called str. Let's try it out.

**Print out the structure of your table**

```{r}
str(my_data)
```


Whenever you are going to use a library that is not native to R, you have to load it into the R environment after you install it. In other words, you need to install once only, but to use it, you must load it into R for every new session.

**Load the "readxl" library into the R environment.**

```{r}
library(readxl)
```


Now that we have our library and its functions ready, we can move on to actually reading the file. In readxl, there is a function called read_excel, which does all the work for us. You can use it like this:

**Read data from the XLS file and attribute the table to the my_excel_data variable.**

```{r}
my_excel_data <- read_excel("movies-db.xls")
```


Since my_excel_data is now a dataframe in R, much like the one we created out of the CSV file, all of the native R functions can be applied to it, like head and str.

Prints out the structure of your table.

Tells you how many rows and columns there are, and the names and type of each column.

**This should be the very same as the other table we created, as they are the same dataset.**

```{r}
str(my_excel_data)
```



## ACCESSING ROWS AND COLUMNS

Whenever we use functions to read tabular data in R, the default method of structuring this data in the R environment is using Data Frames -- R's primary data structure. Data Frames are extremely versatile, and R presents us many options to manipulate them.


Suppose we want to access the "name" column of our dataset. We can directly reference the column name on our data frame to retrieve this data, like this:

**Retrieve a subset of the data frame consisting of the "name" columns**

```{r}
my_data['name'] #<< NOT A VECTOR
```


Another way to do this is by using the $ notation which at the output will provide a vector:

**Retrieve the data for the "name" column in the data frame.**

```{r}
my_data$name # << WILL GIVE A VECTOR
```


**You can also do the same thing using double square brackets, to get a vector of names column.**

```{r}
my_data[["name"]] # << WILL GIVE A VECTOR
```


Similarly, any particular row of the dataset can also be accessed. For example, to get the first row of the dataset with all column values, we can use:

**Retrieve the first row of the data frame.**

```{r}
my_data[1,]
```


We can specify more than one column or row by using c, the concatenate function. By using c to concatenate a list of elements, we tell R that we want these observations out of the data frame. Let's try it out.

**Retrieve the first row of the data frame, but only the "name" and "length_min" columns.**

```{r}
my_data[1, c("name","length_min")]
```



## Accessing Built-in Datasets in R

**Displays a list of the inbuilt datasets. Opens in a new "window".**
```{r}
data()
```


Thankfully, R has documentation present for each inbuilt dataset. You can take a look at that by using the help function.

**Opens up the documentation for the inbuilt "women" dataset.**

```{r}
help(women)
```


Since the datasets listed are inbuilt, you do not need to import or load them to use them. If you reference them by their name, R already has the data frame ready.
```{r}
women
```

