---
title: "Webscraping in R"
author: "Edgar Orong"
date: "2023-05-11"
output: html_document
---

## Webscraping in R


### The `revest` library

**The rvest package is a popular web scraping package for R. After rvest reads an HTML page, you can use the tag names to find the child nodes of the current node.**
```{r}
library(rvest)
```


**We also need to import httr library to get some HTML pages by sending HTTP GET request**
```{r}
library(httr)
```


First let's warm-up by reading HTML from the following character variable simple_html_text

**A simple HTML document in a character variable**
```{r}
simple_html_text <- "
<html>
    <body>
        <p>This is a test html page</p>
    </body>
</html>"
```


Then use the read_html function to create the HTML root node, i.e., the html node by loading the simple_html_text

**Create the root html node by reading the simple HTML string**
```{r}
root_node <- read_html(simple_html_text)
root_node
```


**You can also check the type of root_node**
```{r}
class(root_node)
```


You can see the class is xml_node because rvest load HTML pages and convert them using XML format internally. XML has similar parent-child tree structure but more suitable for storing and tranporting data than HTML.

**Next, let's try to create a HTML node by loading a remote HTML page given a URL**
```{r}
ibm_html_node <- read_html("http://www.ibm.com")
ibm_html_node
```


Sometimes you want to download some HTML pages and analyze them offline, you could use download.file to do so:

**download the R home page and save it to an HTML file locally called r.html**
```{r}
download.file('https://www.r-project.org', destfile = 'r.html')
```


**Create a html_node from the local r.html file**
```{r}
html_node <- read_html('r.html')
html_node

```


Now you know how to read an HTML page from a character variable, a URL, or a local HTML file. 

**Next let's see how to parse and extract data from a specific node(s) starting from the root node**

```{r}
simple_html_text <- "
<html>
    <body>
        <p style=\"color:red;\">This is a test html page</p>
    </body>
</html>"

root_node <- read_html(simple_html_text)
root_node
```


Get the <body> node by using its parent node <html>

**Get the child <body> node from <html> root_node**
```{r}
body_node <- html_node(root_node, "body")
body_node
#You can see it has a child node paragraph <p>
# Let's get the content of the <p>
# Get the child <p> node from its <body> node
p_node <- html_node(body_node, "p")
p_content<-html_text(p_node)
p_content
#The <p> node also has style attribute with value color:red;, which means we want the browser to render its text using red color. To get an attribute of a node, we can use a function called html_attr("attribute name")
# Use the p_node as the first argument to get its attribute
style_attr <- html_attr(p_node, "style")
style_attr
```


In the code cell below, the downloaded r.html file (from https://www.r-project.org) has an <img> node representing an image URL to R logo image (a relative path on its web server), let's try to find the image URL and download it.


You need to paste the relative path in <img> with the the https://www.r-project.org to get the full URL of the image, and use the GET function to request the image as bytes in the response

**Write your code below. Don't forget to press Shift+Enter to execute the cell**
```{r}
url <- 'https://www.r-project.org'
html_node <- read_html('r.html')
```


**Get the image node using its root node**
```{r}
img_node <- html_node(html_node, "img")
# Get the "src" attribute of img node, representing the image location
img_relative_path <- html_attr(img_node, "src")
img_relative_path
```


**Paste img_relative_path with 'https://www.r-project.org'**
```{r}
image_path <- paste(url, img_relative_path, sep="")
```

**Use GET request to get the image**
```{r}
image_response<-GET(image_path)
```


Then use writeBin() function to save the returned bytes into an image file.

**Parse the body from the response as bytes**
```{r}
image <- content(image_response, "raw")
```

**Write the bytes to a png file**
```{r}
writeBin(image, "r.png")
```

Now, from the file list on the left, you should be able to find a saved r.png file.


In HTML, many tabluar data are stored in <table> nodes. Thus, it is very important to be able to extract data from <table> nodes and preferably convert them into R data frames.


Below is a sample HTML page contains a color table showing the supported HTML colors, and we want to load it as a R data frame so we can analyze it using data frame-related operations.

```{r}
table_url <- "https://cf-courses-data.s3.us.cloud-object-storage.appdomain.cloud/IBM-DA0321EN-SkillsNetwork/labs/datasets/HTMLColorCodes.html"

```

**Like other HTML nodes, let's first get the <table> node using html_node function**
```{r}
root_node <-read_html(table_url)
table_node <- html_node(root_node, "table")
table_node
```


You can see the table node in a messy HTML format. Fortunately, you dont need to parse it by yourself, rvest provides a handy function called html_table() to convert <table> node into R dataframe

**Extract content from table_node and convert the data into a dataframe**
```{r}
color_data_frame <- html_table(table_node)
head(color_data_frame)
```


**But you could see the table headers were parsed as the first row, no worries, we could manually fix that**
```{r}
names(color_data_frame)
```


**Convert the first row as column names**
```{r}
names(color_data_frame) <- as.matrix(color_data_frame[1, ])
```

**Then remove the first row**
```{r}
data_frame <- color_data_frame[-1, ]
```

```{r}
head(data_frame)
names(color_data_frame)
```

That's it for webscraping in R, there is a lot more you can do check out rvest.
